What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-08 android-stacks.tex
Unifying the terminology.

Moreover, some languages use exceptions to represent coding errors
(out-of-bounds array index, division-by-zero, access to a null). In these cases
exceptions are implicitly signaled by the runtime environment when such
condition happens. One of such implicit exceptions (caused by data conversion
from a 64-bit floating point to a 16-bit signed integer) was the responsible for
the famous failure on Ariane 5's first test flight~\cite{lions1996ariane} ---
leading to the rocket self-destructing 37 seconds after launch and a loss of 500
million dollars. Besides Ariane 5, several applications crash everyday due to
uncaught exceptions~\cite{jo2004uncaught}. The consequences of not careful
exception usage may have the opposite effect of its initial intention of
improving system robustness.

@@

  \item  A multitude undocumented runtime exceptions signaled by third party
    code (i.e., libraries and Android platform).

@@

\subsection{Error handling in programming languages}

@@

\subsubsection{Exception Throwing and Handling}

@@

\subsection{Exception Handling Best Practices}
\label{sec:best}

%\todo{This section feels a bit disconnected.}
Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. 
% To guide our investigation, we compile the guidelines
A compiled list of the guidelines related to exception types and exception propagation in Java the following:
\footnote{We also compiled guidelines related to exception handling but they are
out of the scope of this paper, since they do not relate to information
available on stack traces.}:

@@

\textbf{IV-Do not repeatedly re-throw exceptions, handle exceptions as close as
possible to the problem}~\cite{wirfs2006toward}. 
%Programs that frequently throw exceptions lose in performance ~\cite{wirfs2006toward,gosling2000java}.
%Moreover, 
Close to the place that the exception was throw the caller usually has
enough contextual information to perform  corrective actions. When the exception
is propagated far away from the method where it was signaled (i.e. further away
from the problem) it can be difficult to take meaningful recovery actions.
todo{RSC:this practice was not well explored, can be removed if we need space.}

%\noindent\emph{Exception Documentation}
\textbf{V-Document all exceptions thrown by a
method}~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to coding
errors} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

@@

Coding errors are the most common root causes for most types of signalers. We
could observe that the NullPointerException is still the exception with higher
number of occurrences. The NullPoiterExceptions are mainly signaled inside
Android platform and inside application code, although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus if it is a good or bad practice to re-throw
NullPointerException. Some prefer to encapsulate such an exception on
IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
should not be wrapped.

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. If we filter the exceptions implicitly signaled by 
JVM (due to coding erros) the number of runtime exceptions thrown by third party code is XXX 
signaled by Android and YYY signaled by library code.

@@

Manually inspecting 10% of each set we could also observe that none of the methods throwing 
runtime exceptions documented the exceptions on the method signature (as suggested by Boch
 ~\cite{}) nor included a comment in Javadoc. This result is aligned with the results of other 
study conducted by from Cabral and Marcques ~\cite{} which  observed that 
runtime exceptions in .Net programs are most often not documented.

@@

This analysis also revealed unexpected wrappings such as: Error wrapping a Checked 
exception; a Runtime wrapping and Error; and an Error wrapping a Runtime.
Java is the only language that provides a hybrid exception handling mechanism
which offers  different types to represent different exception behaviors (i.e., error, 
runtime and checked) (see Section ~\ref{extypes}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to coding errors from which the developer is not expected to recover
from.

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5]{only-wrap-histogram.pdf}
  \end{center}
  \caption{Stack sizes (top) and wrappings (bottom) for Java (left) and Android (right) stack traces}
  \label{fig:sizewrapbhists}
\end{figure}

We could also observe that in 205 stacks found in Android repository the exception was wrapped 
more then once. Figure ~\ref{sizewrapbhists}  illustrates the number of wrappings per stack, both
 in Java and Android studied repositories. One interesting wrapping found in ones Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

Since there is no way of enforcing such conventions during program development,
we could observe the stack trace analysis combined with types these three types
of exceptions were used interchangeably. This interchangeability makes the
behaviour of the exception handling code more complex and less reliable. For
instance, considering a situation were an OutOfMemoryError (a situation that
should not be handled) was wrapped in a checked exception (a situation that must
he handled), in this case this wrapping is telling to the caller that it should
handled.

>>>>

Moreover, some languages use exceptions to represent programming mistakes
(out-of-bounds array index, division-by-zero, access to a null). In these cases
exceptions are implicitly signaled by the runtime environment when such
condition happens. One of such implicit exceptions (caused by data conversion
from a 64-bit floating point to a 16-bit signed integer) was the responsible for
the famous failure on Ariane 5's first test flight~\cite{lions1996ariane} ---
leading to the rocket self-destructing 37 seconds after launch and a loss of 500
million dollars. Besides Ariane 5, several applications crash everyday due to
uncaught exceptions~\cite{jo2004uncaught}. The consequences of not careful
exception usage may have the opposite effect of its initial intention of
improving system robustness.

@@

  \item  A multitude of programming mistakes as the main causes of stack traces.

  \item Undocumented runtime exceptions signaled by third party
    code (i.e., libraries and Android platform).

@@

\subsection{Error Handling in Programming Languages}

@@

%\subsubsection{Exception Throwing and Handling}

@@

\subsection{Exception Handling Best Practices in Java}
\label{sec:best}

%\todo{This section feels a bit disconnected.}
Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. 
% To guide our investigation, we compile the guidelines
A compiled list of the guidelines related to exception types and exception propagation in Java the following
\footnote{We also compiled guidelines related to exception handling but they are
out of the scope of this paper.}:

@@

%\textbf{IV-Do not repeatedly re-throw exceptions, handle exceptions as close as
%possible to the problem}~\cite{wirfs2006toward}. 
%Programs that frequently throw exceptions lose in performance ~\cite{wirfs2006toward,gosling2000java}.
%Moreover, 
%Close to the place that the exception was throw the caller usually has
%enough contextual information to perform  corrective actions. When the exception
%is propagated far away from the method where it was signaled (i.e. further away
%#from the problem) it can be difficult to take meaningful recovery actions.
%*~\todo{RSC:this practice was not well explored, can be removed if we need space.}

%\noindent\emph{Exception Documentation}
\textbf{IV-Document all exceptions thrown by a
method}~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to programming
mistakes} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

@@

Programming mistakes are the most common root causes for most types of signalers. We
could observe that the NullPointerException is still the exception with higher
number of occurrences. The NullPoiterExceptions are mainly signaled inside
Android platform and inside application code, although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus if it is a good or bad practice to re-throw
NullPointerException. Some prefer to encapsulate such an exception on
IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
should not be wrapped.

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. If we filter the exceptions implicitly signaled by 
JVM (due to programming mistakes) the number of runtime exceptions thrown by third party code is XXX 
signaled by Android and YYY signaled by library code.

@@

Manually inspecting 10% of each set we could also observe that none of the methods throwing 
runtime exceptions documented the exceptions on the method signature (as suggested by Bloch
 ~\cite{bloch2008effective}) nor included a comment in Javadoc. This result is aligned with the results of other 
study conducted by from Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

This analysis also revealed unexpected wrappings such as: Error wrapping a Checked 
exception; a Runtime wrapping and Error; and an Error wrapping a Runtime.
Java is the only language that provides a hybrid exception handling mechanism
which offers  different types to represent different exception behaviors (i.e., error, 
runtime and checked) (see Section ~\ref{extypes}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to programming mistakes from which the developer is not expected to recover
from.

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5]{only-wrap-histogram.pdf}
  \end{center}
  \caption{Number of  stack traces wrappings for Java and Android repositories. }
  \label{fig:sizewrapbhists}
\end{figure}

IMPORTANT:  ~\todo{We need to update this data because it is still including PulWifi. Gerogios I sent the filtered data to you by email, and saved on file java-new-stack-sizes.csv in this repository.}

We could also observe that in 205 stacks found in Android repository the exception was wrapped 
more then once. Figure ~\ref{sizewrapbhists}  illustrates the number of wrappings per stack, both
 in Java and Android studied repositories. One interesting multiple-wrapping found in Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

Since there is no way of enforcing Java exception type conventions during program development,
we could observe the stack trace analysis combined with types these three types
of exceptions were used interchangeably. This interchangeability makes the
behaviour of the exception handling code more complex and less reliable. For
instance, considering a situation were an OutOfMemoryError (a situation that
should not be handled) was wrapped in a checked exception (a situation that must
he handled), in this case this wrapping is telling to the caller that it should
handled.

%\section{Discussions}
